\documentclass{beamer}
\usepackage{amsfonts,amsmath,oldgerm, mathtools}
\usepackage{bm}
\usepackage{tikz}
\usepackage{booktabs}
\usepackage{hyperref}
\usetikzlibrary{calc, positioning, arrows.meta, overlay-beamer-styles, shapes.misc, decorations.pathreplacing, backgrounds}
\usetheme{dmpisa}

\renewcommand{\thefootnote}{\arabic{footnote}} % Use numeric footnote markers

\newcommand{\testcolor}[1]{\colorbox{#1}{\textcolor{#1}{test}}~\texttt{#1}}
\usefonttheme[onlymath]{serif}
\titlebackground*{assets/background} % Assicurati che il path sia corretto
\newcommand{\hrefcol}[2]{\textcolor{cyan}{\href{#1}{#2}}}

\title{Efficient Succinct Data Structures on Directed Acyclic Graphs}
\course{Tesi Triennale in Matematica}
\author{\href{mailto:l.lombardo11@studenti.unipi.it}{Luca Lombardo}} % Modifica email se necessario
\date{9 Maggio 2025}

\begin{document}

% --- SLIDE 1: Title Slide ---
\maketitle

% \section{Succinct Data structures}
\include{sections/introduction}

% \section{Rank \& Select}

\include{sections/rank_select/bitvectors}

% \subsection{Strings}

\include{sections/rank_select/wt}

% \subsection{Degenerate Strings}

\include{sections/rank_select/degenerate}

% \section{Path Queries on DAGs}
% \subsection{Path Queries}
\include{sections/succint_dag/dag_path}

% \subsection{Succinct Structure and Queries}
\include{sections/succint_dag/succint}

% \subsection{Querying}
% \include{sections/succint_dag/query}

% \subsection{Compression}
\include{sections/succint_dag/compression}

% \subsection{Achieving Sub-Entropy Space for Path Queries}
\include{sections/succint_dag/entropy}

% \section{Conclusion and Next Steps}
\include{sections/conclusion}
\backmatter

% QUESTIONS?
\begin{frame}[noframenumbering]{Worst-Case $\mathcal{O}$-Set Size: Is Exponential Growth Possible?}
    \framesubtitle{Understanding the $\mathcal{O}$-set Size}

    \begin{block}{Exponential Growth Can Occur}
        The cardinality of an $\mathcal{O}$-set, $|\mathcal{O}_v|$, is not generally bounded by a polynomial in the number of vertices $|V|$. It can grow exponentially.
    \end{block}

    \begin{block}{Underlying Reason: Path Count}
        The number of distinct paths from a source $s$ to a vertex $v$, denoted $|Path(s, v)|$, can itself be exponential in certain DAG structures. Since $|\mathcal{O}_v| \le |Path(s, v)|$, the potential for exponential size exists.
    \end{block}

    \pause % Optional pause

    \begin{alertblock}{Key Condition for Exponential Growth}
        The exponential potential is realized if the vertex weights $w(v)$ are assigned such that distinct paths $P_1 \neq P_2$ almost always lead to distinct cumulative weights $W(P_1) \neq W(P_2)$.
    \end{alertblock}

\end{frame}

\begin{frame}[noframenumbering]{Achieving Exponential $\mathcal{O}$-Set Size}
    \framesubtitle{A Strategy for Path Weight Uniqueness}

    Start with a DAG structure that naturally admits an exponential number of paths between two nodes. An example is a layered graph with multiple choices at each layer transition.

    \begin{alertblock}{Strategic Weight Assignment}
        Assign vertex weights $w(v)$ carefully to ensure path weight uniqueness.
        \[ w(v) = 2^k \quad \text{(using a unique exponent } k \text{ for each node)} \]
    \end{alertblock}

    \pause

    \begin{block}{Mechanism: Unique Binary Representation}
        With power-of-2 weights, the cumulative path weight $W(P) = \sum_{v \in P \setminus \{s\}} w(v)$ becomes a sum of distinct powers of 2.
        Due to the uniqueness of binary representation, different sets of nodes (i.e., different paths) produce different sums.
        Therefore, $|Path(s, v)|$ distinct paths yield $|\mathcal{O}_v| = |Path(s, v)|$ distinct weights.
    \end{block}

\end{frame}

\end{document}
